---
title: "Data Application"
author: "Jun Lu"
date: "2024-05-01"
output: html_document
---

# Load Data
```{r}
library(tidyverse)
library(MCMCvis)
load("./clean_data/binary_5years.Rdata")
load("./clean_data/ttedata.Rdata")
source("f.R")
source("./nimble_code/tte.R")
```

# Plot the data
Issue 1: 890 subjects only have baseline biomarker measurements among 1251 objects (keep or remove them)
```{r}
tte_data %>% 
  ggplot(aes(y = CSFABETA, x = biomarker_days, color = as.factor(Event))) + 
  geom_line(aes(group = NACCID)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_grid(.~as.factor(Event)) +
  theme_bw()
```

# Check number of observations
Obs of biomarkers: Range from 1 to 6
extreme large alpha in this context
Issue 2: The informative cluster size is for survival or binary outcome not for longitudinal mediation. 

mean/median
longitudinal model with cluster size into the model
Disease situation
Poisson
Survival (each n)

informative 

N

(p-value 0.06171) not very significant
(Group with event has more observations)
```{r}
num_obs = tte_data %>%
  arrange(NACCID, biomarker_days) %>% 
  group_by(NACCID, Event) %>% 
  summarise(n = n(),
            mean = first(CSFABETA))

table(num_obs$n)
one.way = aov(mean ~ n, data = num_obs)
summary(one.way)

t.test(n ~ Event, data = num_obs)
num_obs %>% group_by(Event) %>% 
  summarise(n = mean(n))
```


# Check mediation analysis without repeated measures
```{r}
library(mediation)
library(survival)
new_tte_data = tte_data %>% 
  arrange(NACCID, biomarker_days) %>% 
  group_by(NACCID, Event, SEX, EDUC, APOE, age_baseline, Time) %>% 
  summarise(CSFABETA = mean(CSFABETA))

# 48.3%

mod.y <- survreg(Surv(Time, Event) ~ CSFABETA + APOE, data = new_tte_data)
mod.m <- lm(CSFABETA ~ APOE, data = new_tte_data)


med.m <- mediate(mod.m, mod.y, treat = "APOE", mediator = "CSFABETA")
summary(med.m)
summary(mod.y)
summary(mod.m)
```


# Run Model without confounders 
```{r}
d_long = tte_data %>% 
  arrange(NACCID) %>% 
  dplyr::select(NACCID, Event, CSFABETA, APOE, biomarker_days, Time, age_baseline, 
         EDUC, SEX) %>% 
  rename(M = CSFABETA,
         time = biomarker_days,
         A = APOE,
         ID = NACCID,
         ob_t = Time,
         delta = Event) %>% 
  group_by(ID) %>% 
  mutate(obs_n = n()) %>% 
  mutate(SEX = as.numeric(as.factor(SEX)) - 1,
         A = as.numeric(as.factor(A)) - 1)

censured = 1-d_long$delta
t = d_long$ob_t
cent = d_long$ob_t
t[censured==1] = NA
cent[censured ==0] = Inf

d_subject = d_long %>% dplyr::select(-time, -M) %>% distinct()
n = nrow(d_subject)
X1 = cbind(rep(1, n), d_subject$A)

X2 = X1
obs_n = d_subject$obs_n
time = d_long$time
s = rep(0, n)
e = rep(0, n)

# Some issue with writing loop through longitidinal M in NIMBLE
s[1] = 1
for (i in 1:n) {
  e[i] = s[i] + obs_n[i]-1
  if(i != n) {
    s[i+1] = e[i]+1
  }
}

Consts <- list(N1 = n,
               X = X1,
               Q = 2,
               P = 2,
               K = 3,
               s = s,
               e = e,
               Time = d_long$time
               )

Data <- list(M = d_long$M, O = obs_n - 1, censured = censured,
             t = t, cent = cent)

intital = initial_values(d_long, 3)

Inits <- list(L = intital[[1]],
                gamma = intital[[2]],
                b0 = intital[[3]],
                sigma_b = intital[[4]],
                sigma_M = intital[[5]], 
                beta = intital[[6]],
                phi = intital[[7]],
                alpha2 = intital[[8]],
                alpha = intital[[9]],
                theta = intital[[10]])

n.burnin = 10000
Rmodel <- nimbleModel(code, Consts, Data, Inits)
conf <- configureMCMC(Rmodel, print = FALSE, monitors = c('gamma', 'phi'))
conf$removeSamplers('logDens')   
conf$addSampler(target = 'logDens', type = 'sumLogPostDens')
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
    
samples <- runMCMC(mcmc = Cmcmc, 
                   nchains = 1,
                   niter = n.burnin + 500,
                   nburnin = n.burnin,
                   inits = Inits
                   )
    
    MCMCsummary(object = samples, round = 2)
    
    
    list = Map(mediation_effect2, 
               samples[,1],samples[,2], samples[,3], samples[,4],
               samples[,5], samples[,6], samples[,7],samples[,8], 
               samples[,9], samples[,10])
    
    re = do.call(rbind, list)
    colnames(re) = c("NDE", "NIE", "TE", "PROP")
    mean = colMeans(re)
    CI_lower = apply(re, 2, quantile, probs = 0.025)
    CI_upper = apply(re, 2, quantile, probs = 0.975)
    c(mean, CI_lower, CI_upper)

```

# Run Model with confounders 
```{r}
d_long = tte_data %>% 
  arrange(NACCID) %>% 
  select(NACCID, Event, CSFABETA, APOE, biomarker_days, Time, age_baseline, 
         EDUC, SEX) %>% 
  rename(M = CSFABETA,
         time = biomarker_days,
         A = APOE,
         ID = NACCID,
         ob_t = Time,
         delta = Event) %>% 
  group_by(ID) %>% 
  mutate(obs_n = n()) %>% 
  mutate(SEX = as.numeric(as.factor(SEX)) - 1,
         A = as.numeric(as.factor(A)) - 1)

censured = 1-d_long$delta
t = d_long$ob_t
cent = d_long$ob_t
t[censured==1] = NA
cent[censured ==0] = Inf

d_subject = d_long %>% select(-time, -M) %>% distinct()
n = nrow(d_subject)
X1 = cbind(rep(1, n), d_subject$A, d_subject$age_baseline, d_subject$EDUC, d_subject$SEX)

X2 = X1
obs_n = d_subject$obs_n
time = d_long$time
s = rep(0, n)
e = rep(0, n)

# Some issue with writing loop through longitidinal M in NIMBLE
s[1] = 1
for (i in 1:n) {
  e[i] = s[i] + obs_n[i]-1
  if(i != n) {
    s[i+1] = e[i]+1
  }
}

Consts <- list(N1 = n,
               X = X1,
               Q = 2,
               P = 5,
               K = 3,
               s = s,
               e = e,
               Time = d_long$time
               )

Data <- list(M = d_long$M, O = obs_n - 1, censured = censured,
             t = t, cent = cent)
intital = initial_values2(d_long, 3)

Inits <- list(L = intital[[1]],
                gamma = intital[[2]],
                b0 = intital[[3]],
                sigma_b = intital[[4]],
                sigma_M = intital[[5]], 
                beta = intital[[6]],
                phi = intital[[7]],
                alpha2 = intital[[8]],
                alpha = intital[[9]],
                theta = intital[[10]])

n.burnin = 50000
Rmodel <- nimbleModel(code, Consts, Data, Inits)
conf <- configureMCMC(Rmodel, print = FALSE, monitors = c('gamma', 'phi'))
conf$removeSamplers('logDens')   
conf$addSampler(target = 'logDens', type = 'sumLogPostDens')
Rmcmc <- buildMCMC(conf)
Cmodel <- compileNimble(Rmodel)
Cmcmc <- compileNimble(Rmcmc, project = Rmodel)
    
samples <- runMCMC(mcmc = Cmcmc, 
                   nchains = 1,
                   niter = n.burnin + 1000,
                   nburnin = n.burnin,
                   inits = Inits
                   )
    
    MCMCsummary(object = samples, round = 2)
    re = matrix(NA, nrow = 500, ncol = 4)
    for (i in 1:500) {
      gamma = matrix(samples[i,1:10], nrow = 2, ncol = 5)
      gamma = rbind(rep(0,5), gamma)
      phi = matrix(samples[i,11:25], nrow = 3, ncol = 5)
      re[i,] = mediation_effect(gamma, phi, X1)
    }
    
    colnames(re) = c("NDE", "NIE", "TE", "PROP")
    mean = colMeans(re)
    CI_lower = apply(re, 2, quantile, probs = 0.025)
    CI_upper = apply(re, 2, quantile, probs = 0.975)
    c(mean, CI_lower, CI_upper)

```





